---
title: "hIBD hypothalamus steroid sac adults June 2025"
author: "Annie Ciernia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of microglia morphlogy from brains taken 24hrs after last RGFP/LPS injection

#Iba1 staining in both male and female mice used #C57Bl6J background

```{r load libraries}

library(cowplot)
library(stringr)
library(ggplot2)
library(tidyverse)
library(plotrix)
library(rstatix)
library(ggpubr)
library(nlme)
library(emmeans)

library(MicrogliaMorphologyR)
library(factoextra)
library(ppclust)
set.seed(1)

```

feb 21, 2025

```{r Load all data HC}

setwd("/Users/aciernia/Sync/CierniaLabMembers/AnnieCiernia/Experiments/SCFA_Invivo2025/morphology_all")
fraclac.dir <- "/Volumes/NINC/CierniaLab/Clianta/SCFA4/SCFA4\ Microglia\ Morphology/FracLac/20250804084752"
skeleton.dir <- "/Volumes/NINC/CierniaLab/Clianta/SCFA4/SCFA4\ Microglia\ Morphology/SkeletonResults"


# these steps may be very time-intensive, depending on how many cells you are analyzing (i.e., on the order of 1000s of cells). 
fraclac <- fraclac_tidying(fraclac.dir) 
skeleton <- skeleton_tidying(skeleton.dir)

save(fraclac, skeleton, file = "tmp_InputData.Rdata")

#fix mouse 3 names
#skeleton$Name <- gsub(" mouse3_","mouse3_",skeleton$Name)

#check names match
     
skeleton <- skeleton %>% distinct()
data <- merge_data(fraclac, skeleton)
finaldata_nonHC <- metadata_columns(data, c("MouseID","Sex","Region", "Antibody","Cohort","Subregion"), sep="_")

#read in metadata
metadata <- read.csv("SCFA4_metadata_final.csv")

# Extract the last number
metadata$cohort <- str_extract(metadata$Cohort, "\\d+(?!.*\\d)")

metadata$Sex <- toupper(metadata$Sex)
  
metadata$MouseID <- paste("Mouse",metadata$ID,metadata$Sex, metadata$cohort, sep="")

#fix to include sex and cohort
finaldata_nonHC$MouseID <- paste(finaldata_nonHC$MouseID, finaldata_nonHC$Sex, sep="")


metadata$MouseID %in% unique(finaldata_nonHC$MouseID)
unique(finaldata_nonHC$MouseID) %in% metadata$MouseID

missing_ids <- metadata$MouseID[!(metadata$MouseID %in% unique(finaldata_nonHC$MouseID))]

#mouse9M2 is missing

finaldata_nonHC <- merge(metadata, finaldata_nonHC, by="MouseID")
finaldata_nonHC <- distinct(finaldata_nonHC)



save(finaldata_nonHC, file = "nonHC_InputData_tmp.Rdata")



```



```{r}
#read in HC datat

load("/Users/aciernia/Sync/CierniaLabMembers/AnnieCiernia/Experiments/SCFA_Invivo2025/morphology_all/HC_InputData.Rdata")


skeleton <- skeleton %>% distinct()
data <- merge_data(fraclac, skeleton)
finaldata_HC <- metadata_columns(data, c("MouseID","Sex","Region", "Antibody","Cohort","Subregion"), sep="_")

#read in metadata
metadata <- read.csv("SCFA4_metadata_final.csv")

# Extract the last number
metadata$cohort <- str_extract(metadata$Cohort, "\\d+(?!.*\\d)")

metadata$Sex <- toupper(metadata$Sex)

metadata$MouseID <- paste("Mouse",metadata$ID,metadata$Sex, metadata$cohort, sep="")

#fix to include sex and cohort
finaldata_HC$MouseID <- paste(finaldata_HC$MouseID, finaldata_HC$Sex, sep="")


metadata$MouseID %in% unique(finaldata_HC$MouseID)
unique(finaldata_HC$MouseID) %in% metadata$MouseID

missing_ids <- metadata$MouseID[!(metadata$MouseID %in% unique(finaldata_HC$MouseID))]

finaldata_HC <- merge(metadata, finaldata_HC, by="MouseID")
finaldata_HC <- distinct(finaldata_HC)



save(finaldata_HC, file = "HC_InputData.Rdata")



```


```{r}
#combine datasets


finaldata_all <- rbind(finaldata_HC, finaldata_nonHC)


unique(finaldata_all$Sex.x)

unique(finaldata_all$Condition)

unique(finaldata_all$Injection)

unique(finaldata_all$Cohort.x)


unique(finaldata_all$Region)

unique(finaldata_all$Subregion)


#replace values in region with subregion for DH

finaldata_all <- finaldata_all %>%
  mutate(Region = ifelse(Region == "DH", Subregion, Region))

unique(finaldata_all$Region)

```




```{r gather data and normalize}
# gather your numerical morphology data into one column ('measure') which contains the feature name, and another column ('value') which contains measured values


#print(finaldata)

# transform your data in appropriate manner for downstream analyses
data_logtransformed <- transform_log(finaldata_all, 1, start=23, end=ncol(finaldata_all)) 

```



```{r}


library(dplyr)

# Step 1: Apply IQR filter (use your updated IQR logic)
cols_to_check <- 23:49
outlier_indices <- list()

for (col in cols_to_check) {
  x <- data_logtransformed[[col]]
  
  # Skip non-numeric
  if (!is.numeric(x)) next

  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1

  lower_bound <- Q1 - 3 * IQR_val  # use 3.5 for leniency
  upper_bound <- Q3 + 3 * IQR_val

  outlier_indices[[col]] <- which(x < lower_bound | x > upper_bound)
}

# Combine and filter
all_outliers <- unique(unlist(outlier_indices))
df_filtered <- data_logtransformed[-all_outliers, ]

# Step 2: Safely coerce columns 7:33 to numeric
df_filtered[, 23:49] <- df_filtered[,23:49] %>%
  mutate(across(everything(), ~ suppressWarnings(as.numeric(as.character(.)))))

dim(df_filtered)

```

```{r}


constant_cols <- which(apply(df_filtered[,c(23:49)], 2, function(x) var(x, na.rm = TRUE) == 0))
names(df_filtered)[constant_cols]

unique(data_logtransformed$`# of junctions`)
unique(df_filtered$`# of junctions`)


# Step 1: Split metadata and numeric features
metadata <- df_filtered[, 1:23]
features <- df_filtered[, 24:ncol(df_filtered)]

# Step 2: Clean feature columns
features_clean <- features %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select_if(~ var(., na.rm = TRUE) != 0) %>%
  mutate(across(everything(), ~ ifelse(is.finite(.), ., NA))) %>%
  na.omit()

# Step 3: Match metadata rows (only those that survived na.omit)
metadata_clean <- metadata[rownames(features_clean), ]

# Step 4: Combine cleaned metadata + features
clean_data <- cbind(metadata_clean, features_clean)

dim(clean_data)

```



```{r counts}



#save table of number of cells/mouse/subregion
t <- table(clean_data$Region, clean_data$MouseID)
t

#CA2 not enough cells. remove
#mouse 11F2 has very few cells. remove.

clean_data <- clean_data %>% filter(Region != "CA2") 


write.csv(t, "Microglia_PerMousePersubregion.csv")


samples <- clean_data  %>% ungroup() %>% select(1:24) %>% unique()
t <- table(samples$Region, samples$Condition, samples$Injection, samples$Sex.x)
t
write.csv(t, "Mice_per_subregion_condition.csv")


```


## clustering analysis

#Kmeans (no filtering)

```{r PCA and kmeans clustering}


#feature map
pdf("featurecorrelations.pdf", width=10, height=10)

featurecorrelations(clean_data, featurestart=23, featureend=49, rthresh=0.8, pthresh=0.05, title="Correlations across log transformed features")
dev.off()

#featurecorrelations(data_2xLPS_logtransformed, start=15, end=ncol(finaldata2), rthresh=0.8, pthresh=0.05, title="Correlations across log transformed features")

#remove cells with area value missing b/c outliler
#data_logtransformed <- na.omit(data_logtransformed)



#pca
pdf("Elbowplot.pdf", width=6, height=4)
pcadata_elbow(clean_data , featurestart=23, featureend=ncol(clean_data ))
dev.off()

pca_data <- pcadata(clean_data ,featurestart=23, featureend=ncol(clean_data),
                    pc.start=1, pc.end=10)

head(pca_data)

```

```{r feature correlations by PC}

pdf("PCcorrelations.pdf", width=6, height=4)

pcfeaturecorrelations(pca_data, pc.start=1, pc.end=3, 
                      feature.start=33, feature.end=ncol(pca_data), 
                      rthresh=0.75, pthresh=0.05, 
                      title="Correlation between PCs and features")

dev.off()

```




```{r cluster number}
## for k-means clustering: scale PCs 1-3, which together describe ~85% of variability
pca_data_scale <- transform_scale(pca_data, start=1, end=3) 

# scale pca data as input for k-means clustering
kmeans_input <- pca_data_scale[1:3]

# check for optimal number of clusters using wss and silhouette methods
sampling <- kmeans_input[sample(nrow(kmeans_input), 2000),] #sample 5000 random rows for cluster optimization

pdf("nbclust.pdf", width=6, height=4)

fviz_nbclust(sampling, kmeans, method = 'wss', nstart=25, iter.max=50) # 4 clusters

dev.off()


pdf("silhouette.pdf", width=6, height=4)

fviz_nbclust(sampling, kmeans, method = 'silhouette', nstart=25, iter.max=50)
dev.off()

```




```{r}
# cluster and combine with original data
data_kmeans <- kmeans(kmeans_input, centers=4)
data_kmeans

pca_kmeans <- cbind(pca_data[1:3], clean_data, as.data.frame(data_kmeans$cluster))

colnames(pca_kmeans)

colnames(pca_kmeans)[53] <- c("Cluster")

#str(pca_kmeans)

plot <- clusterplots(pca_kmeans, "PC1", "PC2")
plot

pdf("clusterplot.pdf", width=6, height=4)

plot + scale_colour_viridis_d()

dev.off()

```


```{r}
library(plotly)
library(htmlwidgets)


# Ensure Cluster is a factor for discrete coloring
pca_kmeans$Cluster <- as.factor(pca_kmeans$Cluster)

# Choose a manual color palette (can be any valid R color names or hex codes)
cluster_colors <- RColorBrewer::brewer.pal(n = length(unique(pca_kmeans$Cluster)), name = "Set2")

# Plot with manual colors
fig <- plot_ly(
  data = pca_kmeans,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~Cluster,
  colors = cluster_colors,  # use discrete palette
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5, opacity = 0.8)
)

# Save as HTML file
saveWidget(fig, file = "3D_PCA_plot.html", selfcontained = TRUE)




```




```{r}

pdf("clusterheatmap_numbers.pdf", width=6, height=4)
clusterfeatures(pca_kmeans, featurestart=26, featureend =52)
dev.off()

clusterfeatures(pca_kmeans, featurestart=26, featureend =52)


```



#manually label

```{r}

# update cluster labels after looking at color by cluster
data_pca_kmeans  <- pca_kmeans %>% mutate(Cluster = 
                      case_when(Cluster=="1" ~ "ramified",
                                Cluster=="2" ~ "ameboid",
                                Cluster=="3" ~ "rod-like",
                                Cluster=="4" ~ "hypertrophic"))

pdf("clusterheatmap_names.pdf", width=6, height=4)

clusterfeatures(data_pca_kmeans,featurestart=26, featureend =52)

dev.off()


save(data_kmeans, kmeans_input, pca_data, pca_kmeans, pca_data_scale, file= "Data_clusterdata.RData")

clusterfeatures(data_pca_kmeans,featurestart=26, featureend =52)


```

```{r plots by whole HC}
# Quick check of cluster proportions when considering experimental variables of interest

cp2 <- clusterpercentage(pca_kmeans, "Cluster", MouseID, Region, Sex.x, Condition, Injection)


cp2 <- cp2  %>%  mutate(Cluster = 
                           case_when(Cluster=="1" ~ "ramified",
                                Cluster=="2" ~ "ameboid",
                                Cluster=="3" ~ "rod-like",
                                Cluster=="4" ~ "hypertrophic"))


cp2$Condition <- factor(cp2$Condition, levels = c("Saline","Butyrate"  ,  "Propionate", "Acetate"))

cp2$Injection <- factor(cp2$Injection, levels= c("PBS","LPS"))

cp2$Region <- factor(cp2$Region, levels= c("CA1" ,"CA3", "DG","PFC","CC" , "CP"))


library(lemon)
library(cowplot)

pdf("Boplot_subregion_SCFA_Sex_Cluster.pdf", height = 40, width =16)    # create PNG for the heat map       

cbPalette <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7")
p <- cp2 %>%
  ggplot(aes(x = Cluster, y = percentage * 100, fill = Injection)) +
  facet_rep_wrap(~ Region * Sex.x *Condition, repeat.tick.labels = 'bottom', ncol = 4) +
  
  stat_summary(
    geom = "boxplot",
    fun.data = function(x) setNames(quantile(x, c(0.05, 0.25, 0.5, 0.75, 0.95)),
                                    c("ymin", "lower", "middle", "upper", "ymax")),
    position = position_dodge(width = 0.9)
  ) +

  geom_point(aes(group = Injection),
             position = position_dodge(width = 0.9),
             shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +

  scale_fill_manual(values = cbPalette) +
  scale_x_discrete(name = "Cluster") +
  scale_y_continuous(name = "Percentage") +
  ggtitle("K-means clusters") +
  theme_cowplot(font_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

dev.off();

write.csv(cp2,"Mean_Cluster_data.csv")

pdf("Boplot_subregion_SCFA_Injection_Cluster.pdf", height = 30, width =16)    # create PNG for the heat map       

cbPalette <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7")
p <- cp2 %>%
  ggplot(aes(x = Cluster, y = percentage * 100, fill = Condition)) +
  facet_rep_wrap(~ Region * Sex.x *Injection, repeat.tick.labels = 'bottom', ncol = 4) +
  
  stat_summary(
    geom = "boxplot",
    fun.data = function(x) setNames(quantile(x, c(0.05, 0.25, 0.5, 0.75, 0.95)),
                                    c("ymin", "lower", "middle", "upper", "ymax")),
    position = position_dodge(width = 0.9)
  ) +

  geom_point(aes(group = Condition),
             position = position_dodge(width = 0.9),
             shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +

  scale_fill_manual(values = cbPalette) +
  scale_x_discrete(name = "Cluster") +
  scale_y_continuous(name = "Percentage") +
  ggtitle("K-means clusters") +
  theme_cowplot(font_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

dev.off();


p
```



```{r stats analysis for HYP}

# prepare percentages dataset for downstream analysis
stats.input <- cp2
stats.input$Sex <- factor(stats.input$Sex.x)
stats.input$Cluster <- factor(stats.input$Cluster)
stats.input$Subregion <- factor(stats.input$Region)
table(is.na(stats.input))



```


```{r}
#loop for each region

# Get unique subregions
subregions <- unique(stats.input$Subregion)

#CC throws an error
subregions <- c("CA1" ,"CA3", "DG","PFC","CP")


  
  
# Initialize lists to store results
all_anovas <- list()
all_posthoc1 <- list()
all_posthoc2 <- list()

# Loop over each subregion
for (region in subregions) {
  
  # Subset data
  stats_sub <- stats.input[stats.input$Subregion == region, ]
  
  # Run stats model
  stats.testing <- stats_cluster.animal(
    stats_sub,
    "percentage ~ Cluster*Condition*Injection*Sex + (1|MouseID)",  # no 'subregion' term
    "~Condition|Injection|Cluster*Sex", 
    "~Injection|Condition|Cluster*Sex", 
    "sidak"
  )
  
  # Extract and store ANOVA
  anova <- stats.testing[[1]]
  all_anovas[[region]] <- anova
 # write.csv(anova, paste0("ANOVA_", region, "_kmeansclusters.csv"))
  
  # Extract and store posthoc1
  posthoc1 <- stats.testing[[2]]
  posthoc1$correction <- "sidak"
  all_posthoc1[[region]] <- posthoc1
 # write.csv(posthoc1, paste0("Posthoc1_", region, "_Sidak_kmeansclusters.csv"))
  
  # Extract and store posthoc2
  posthoc2 <- stats.testing[[3]]
  posthoc2$correction <- "sidak"
  all_posthoc2[[region]] <- posthoc2
 # write.csv(posthoc2, paste0("Posthoc2_", region, "_Sidak_kmeansclusters.csv"))
}

# Optional: combine and export all results into one table each
combined_anova <- dplyr::bind_rows(all_anovas, .id = "Subregion")
combined_posthoc1 <- dplyr::bind_rows(all_posthoc1, .id = "Subregion")
combined_posthoc2 <- dplyr::bind_rows(all_posthoc2, .id = "Subregion")


#add significance column
combined_anova <- combined_anova  %>%
  mutate(Significance = ifelse(`Pr(>Chisq)` < 0.05, "significant", "ns"))


# Save combined files
write.csv(combined_anova, "Combined_ANOVA_AllSubregions.csv")
write.csv(combined_posthoc1, "Combined_Posthoc1_Sidak_AllSubregions.csv")
write.csv(combined_posthoc2, "Combined_Posthoc2_Sidak_AllSubregions.csv")



```


```{r}
# Required libraries
library(ggplot2)
library(tibble)

# Create a data frame for the legend
legend_df <- tibble::tibble(
  Cluster = factor(c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster 4"),
                   levels = c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster 4")),
  Color = c("#BBCC33", "#44BB99", "#EEDD88", "#EE8866"),
  Label = c("Cluster 1: ramified", 
            "Cluster 2: ameboid", 
            "Cluster 3: rod-like", 
            "Cluster 4: hypertrophic")
)



pdf("Boplot_legend.pdf", height = 6, width =8)    # create PNG for the heat map       

ggplot(legend_df, aes(x = Cluster, y = 1, fill = Cluster)) +
  geom_tile(color = "black") +
  scale_fill_manual(values = legend_df$Color) +
  theme_void() +
  theme(legend.position = "none") +
  geom_text(aes(label = Label), color = "black", hjust = 0.5, vjust = 0.5, size = 5) +
  coord_fixed(ratio = 0.2) +
  ggtitle("Microglial Morphology Clusters Legend") +
  theme(plot.title = element_text(hjust = 0.5, size = 12))

dev.off()

```


```{r color by cluster kmeans clusters}
###############################################################################
#rcolor by cluster
################################################################################


animals <- unique(as.character(stats.input$MouseID))

#no CA2
subregions <- c("CA1" ,"CA3", "DG","PFC","CP")

#for each subsubregion and mouse
for (a in animals) {

#filter for each subsubregion
# isolate out all the cells for your specific image of interest
for (s in subregions) {
  
  colorbycluster <- pca_kmeans %>%
  filter(MouseID == a,  Region == s) 
    
   colorbycluster$filename <- sub("_[^_]*$", "", colorbycluster$UniqueID)
    filename <- unique(colorbycluster$filename)
    
   colorbycluster <- colorbycluster %>% select(c(Cluster, ID.y))

  colnames(colorbycluster) <- c("Cluster","ID")
  
  write.csv(colorbycluster, file= paste(filename,".csv",sep=""))
  
}

}

save.image(file="AllsubregionsAnalysis2025.RData")

```



```{r}
samples <- clean_data  %>% ungroup() %>% select(1:24) %>% unique() %>% 
  filter(Region != "CC") %>% 
  filter(Region != "CA2")


t <- table(samples$Region, samples$Condition, samples$Injection, samples$Sex.x)
t
write.csv(t, "Mice_per_subregion_condition.csv")


samples <- clean_data  %>% ungroup() %>% select(1:17) %>% unique() %>% 
  filter(Region != "CC") %>% 
  filter(Region != "CA2")


t <- table(samples$Region, samples$Condition, samples$Injection, samples$Sex.x)
t

```


```{r}
#graph without LPS
# Quick check of cluster proportions when considering experimental variables of interest

cp2 <- clusterpercentage(pca_kmeans, "Cluster", MouseID, Region, Sex.x, Condition, Injection)


cp2 <- cp2  %>%  mutate(Cluster = 
                           case_when(Cluster=="1" ~ "ramified",
                                Cluster=="2" ~ "ameboid",
                                Cluster=="3" ~ "rod-like",
                                Cluster=="4" ~ "hypertrophic"))


cp2$Condition <- factor(cp2$Condition, levels = c("Saline","Butyrate"  ,  "Propionate", "Acetate"))

cp2 <- cp2 %>% filter(Injection == "PBS")

cp2$Region <- factor(cp2$Region, levels= c("CA1" ,"CA3", "DG","PFC","CC" , "CP"))


library(lemon)
library(cowplot)

pdf("Boplot_subregion_SCFA_Sex_Cluster_noLPS.pdf", height =8, width =12)    # create PNG for the heat map       

cbPalette <- c("#0072B2", "#E69F00", "#009E73", "#CC79A7", "pink","skyblue" )

p <- cp2 %>%
  ggplot(aes(x = Cluster, y = percentage * 100)) +
  facet_rep_wrap(~ Region, repeat.tick.labels = 'bottom', ncol = 3, scales ="free") +

  # Boxplots grouped by Cluster + Condition
  stat_summary(
    aes(group = interaction(Cluster, Condition), fill = Condition),
    geom = "boxplot",
    fun.data = function(x) setNames(
      quantile(x, c(0.05, 0.25, 0.5, 0.75, 0.95)),
      c("ymin", "lower", "middle", "upper", "ymax")
    ),
    position = position_dodge(width = 0.9),
    width = 0.6
  ) +

  # Dots grouped by Cluster + Condition + Sex.x
  geom_point(
    aes(group = interaction(Cluster, Condition), fill = Sex.x),
    position = position_dodge(width = 0.9),
    shape = 21,
    size = 2,
    alpha = 0.7,
    stroke = 0.3
  ) +

  scale_fill_manual(values = cbPalette) +
  scale_x_discrete(name = "Cluster") +
  scale_y_continuous(name = "Percentage") +
  ggtitle("K-means clusters") +
  theme_cowplot(font_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

dev.off()






```


```{r}
#loop for each region only for PBS


stats.input <- cp2


#CC throws an error
subregions <- c("CA1" ,"CA3", "DG","PFC","CP")


  
  
# Initialize lists to store results
all_anovas <- list()
all_posthoc1 <- list()
all_posthoc2 <- list()

# Loop over each subregion
for (region in subregions) {
  
  # Subset data
  stats_sub <- stats.input[stats.input$Region == region, ]
  
  # Run stats model
  stats.testing <- stats_cluster.animal(
    stats_sub,
    "percentage ~ Cluster*Condition*Sex.x + (1|MouseID)",  # no 'subregion' term
    "~Condition|Cluster*Sex.x", 
    "~Condition|Cluster", 
    "sidak"
  )
  
  # Extract and store ANOVA
  anova <- stats.testing[[1]]
  all_anovas[[region]] <- anova
 # write.csv(anova, paste0("ANOVA_", region, "_kmeansclusters.csv"))
  
  # Extract and store posthoc1
  posthoc1 <- stats.testing[[2]]
  posthoc1$correction <- "sidak"
  all_posthoc1[[region]] <- posthoc1
 # write.csv(posthoc1, paste0("Posthoc1_", region, "_Sidak_kmeansclusters.csv"))
  
  # Extract and store posthoc2
  posthoc2 <- stats.testing[[3]]
  posthoc2$correction <- "sidak"
  all_posthoc2[[region]] <- posthoc2
 # write.csv(posthoc2, paste0("Posthoc2_", region, "_Sidak_kmeansclusters.csv"))
}

# Optional: combine and export all results into one table each
combined_anova <- dplyr::bind_rows(all_anovas, .id = "Subregion")
combined_posthoc1 <- dplyr::bind_rows(all_posthoc1, .id = "Subregion")
combined_posthoc2 <- dplyr::bind_rows(all_posthoc2, .id = "Subregion")


#add significance column
combined_anova <- combined_anova  %>%
  mutate(Significance = ifelse(`Pr(>Chisq)` < 0.05, "significant", "ns"))


# Save combined files
write.csv(combined_anova, "Combined_ANOVA_AllSubregions_noLPS.csv")
write.csv(combined_posthoc1, "Combined_Posthoc1_Sidak_AllSubregions_noLPS.csv")
write.csv(combined_posthoc2, "Combined_Posthoc2_Sidak_AllSubregions_noLPS.csv")



```
```{r}
#graph without Butyarate only
# Quick check of cluster proportions when considering experimental variables of interest

cp2 <- clusterpercentage(pca_kmeans, "Cluster", MouseID, Region, Sex.x, Condition, Injection)


cp2 <- cp2  %>%  mutate(Cluster = 
                           case_when(Cluster=="1" ~ "ramified",
                                Cluster=="2" ~ "ameboid",
                                Cluster=="3" ~ "rod-like",
                                Cluster=="4" ~ "hypertrophic")) %>%
  filter(Condition == "Saline"| Condition == "Butyrate")


cp2$Region <- factor(cp2$Region, levels= c("CA1" ,"CA3", "DG","PFC","CC" , "CP"))
cp2$Condition <- factor(cp2$Condition, levels = c("Saline","Butyrate"))
cp2$Injection <- factor(cp2$Injection, levels = c("PBS","LPS"))
cp2$Sex.x <- factor(cp2$Sex.x, levels = c("M","F"))
cp2$Group <- paste(cp2$Condition,cp2$Injection, sep= " ")
cp2$Group <- factor(cp2$Group, levels = c("Saline PBS","Butyrate PBS","Saline LPS","Butyrate LPS"))

library(lemon)
library(cowplot)

pdf("Boplot_subregion_SCFA_Sex_Cluster_butyrateonly.pdf", height =8, width =12)    # create PNG for the heat map       

cbPalette <- c("#0072B2", "#009E73", "#E69F00", "#CC79A7","skyblue", "pink" )

p <- cp2 %>%
  ggplot(aes(x = Cluster, y = percentage * 100), group = Group) +
  facet_rep_wrap(~ Region, repeat.tick.labels = 'bottom', ncol = 3, scales ="free") +

  # Boxplots grouped by Cluster + Condition
  stat_summary(
    aes(group = interaction(Cluster, Group), fill = Group),
    geom = "boxplot",
    fun.data = function(x) setNames(
      quantile(x, c(0.05, 0.25, 0.5, 0.75, 0.95)),
      c("ymin", "lower", "middle", "upper", "ymax")
    ),
    position = position_dodge(width = 0.9),
    width = 0.6
  ) +

  # Dots grouped by Cluster + Condition + Sex.x
  geom_point(
    aes(group = interaction(Cluster, Group), fill = Sex.x),
    position = position_dodge(width = 0.9),
    shape = 21,
    size = 2,
    alpha = 0.7,
    stroke = 0.3
  ) +

  scale_fill_manual(values = cbPalette) +
  scale_x_discrete(name = "Cluster") +
  scale_y_continuous(name = "Percentage") +
  ggtitle("K-means clusters") +
  theme_cowplot(font_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

dev.off()






```

```{r}
#loop for each region only for PBS


stats.input <- cp2


#CC throws an error
subregions <- c("CA1" ,"CA3", "DG","PFC","CP")


  
  
# Initialize lists to store results
all_anovas <- list()
all_posthoc1 <- list()
all_posthoc2 <- list()

# Loop over each subregion
for (region in subregions) {
  
  # Subset data
  stats_sub <- stats.input[stats.input$Region == region, ]
  
  # Run stats model
  stats.testing <- stats_cluster.animal(
    stats_sub,
    "percentage ~ Cluster*Condition*Injection*Sex.x + (1|MouseID)",  # no 'subregion' term
    "~Condition|Injection|Cluster*Sex.x", 
    "~Condition|Injection|Cluster", 
    "sidak"
  )
  
  # Extract and store ANOVA
  anova <- stats.testing[[1]]
  all_anovas[[region]] <- anova
 # write.csv(anova, paste0("ANOVA_", region, "_kmeansclusters.csv"))
  
  # Extract and store posthoc1
  posthoc1 <- stats.testing[[2]]
  posthoc1$correction <- "sidak"
  all_posthoc1[[region]] <- posthoc1
 # write.csv(posthoc1, paste0("Posthoc1_", region, "_Sidak_kmeansclusters.csv"))
  
  # Extract and store posthoc2
  posthoc2 <- stats.testing[[3]]
  posthoc2$correction <- "sidak"
  all_posthoc2[[region]] <- posthoc2
 # write.csv(posthoc2, paste0("Posthoc2_", region, "_Sidak_kmeansclusters.csv"))
}

# Optional: combine and export all results into one table each
combined_anova <- dplyr::bind_rows(all_anovas, .id = "Subregion")
combined_posthoc1 <- dplyr::bind_rows(all_posthoc1, .id = "Subregion")
combined_posthoc2 <- dplyr::bind_rows(all_posthoc2, .id = "Subregion")


#add significance column
combined_anova <- combined_anova  %>%
  mutate(Significance = ifelse(`Pr(>Chisq)` < 0.05, "significant", "ns"))


# Save combined files
write.csv(combined_anova, "Combined_ANOVA_AllSubregions_butyrateonly.csv")
write.csv(combined_posthoc1, "Combined_Posthoc1_Sidak_AllSubregions_butyrateonly.csv")
write.csv(combined_posthoc2, "Combined_Posthoc2_Sidak_AllSubregions_butyrateonly.csv")



```



``````{r}
#graph without Butyarate only
# Quick check of cluster proportions when considering experimental variables of interest

cp2 <- clusterpercentage(pca_kmeans, "Cluster", MouseID, Region, Sex.x, Condition, Injection)


cp2 <- cp2  %>%  mutate(Cluster = 
                           case_when(Cluster=="1" ~ "ramified",
                                Cluster=="2" ~ "ameboid",
                                Cluster=="3" ~ "rod-like",
                                Cluster=="4" ~ "hypertrophic")) %>%
  filter(Condition == "Saline"| Condition == "Butyrate")

cp2 <- cp2 %>% filter(Cluster=="ameboid")

cp2$Region <- factor(cp2$Region, levels= c("CA1" ,"CA3", "DG","PFC","CC" , "CP"))
cp2$Condition <- factor(cp2$Condition, levels = c("Saline","Butyrate"))
cp2$Injection <- factor(cp2$Injection, levels = c("PBS","LPS"))
cp2$Sex.x <- factor(cp2$Sex.x, levels = c("M","F"))
cp2$Group <- paste(cp2$Condition,cp2$Injection, sep= " ")
cp2$Group <- factor(cp2$Group, levels = c("Saline PBS","Butyrate PBS","Saline LPS","Butyrate LPS"))

library(lemon)
library(cowplot)

pdf("Boplot_subregion_SCFA_Sex_Cluster_butyrateonly_ameboidonly.pdf", height =12, width =8)    # create PNG for the heat map       

cbPalette <- c("#0072B2", "#009E73", "#E69F00", "#CC79A7","skyblue", "pink" )

p <- cp2 %>%
  ggplot(aes(x = Injection, y = percentage * 100), group = Group) +
  facet_rep_wrap(~ Region*Sex.x, repeat.tick.labels = 'bottom', ncol = 3) +

  # Boxplots grouped by Cluster + Condition
  stat_summary(
    aes(group = interaction(Cluster, Injection, Condition), fill = Condition),
    geom = "boxplot",
    fun.data = function(x) setNames(
      quantile(x, c(0.05, 0.25, 0.5, 0.75, 0.95)),
      c("ymin", "lower", "middle", "upper", "ymax")
    ),
    position = position_dodge(width = 0.9),
    width = 0.6
  ) +

  # Dots grouped by Cluster + Condition + Sex.x
  geom_point(
    aes(group = interaction(Cluster, Injection, Condition), fill = Sex.x),
    position = position_dodge(width = 0.9),
    shape = 21,
    size = 2,
    alpha = 0.7,
    stroke = 0.3
  ) +

  scale_fill_manual(values = cbPalette) +
  scale_x_discrete(name = "Cluster") +
  scale_y_continuous(name = "Percentage") +
  ggtitle("Ameboid Microglia") +
  theme_cowplot(font_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

dev.off()


``````

```{r}
#loop for each region only for PBS


stats.input <- cp2


#CC throws an error
subregions <- c("CA1" ,"CA3", "DG","PFC","CP")


  
  
# Initialize lists to store results
all_anovas <- list()
all_posthoc1 <- list()
all_posthoc2 <- list()

# Loop over each subregion
for (region in subregions) {
  
  # Subset data
  stats_sub <- stats.input[stats.input$Region == region, ]
  
  # Run stats model
  stats.testing <- stats_cluster.animal(
    stats_sub,
    "percentage ~ Condition*Injection*Sex.x + (1|MouseID)",  # no 'subregion' term
    "~Condition|Injection*Sex.x", 
    "~Condition|Injection", 
    "sidak"
  )
  
  # Extract and store ANOVA
  anova <- stats.testing[[1]]
  all_anovas[[region]] <- anova
 # write.csv(anova, paste0("ANOVA_", region, "_kmeansclusters.csv"))
  
  # Extract and store posthoc1
  posthoc1 <- stats.testing[[2]]
  posthoc1$correction <- "sidak"
  all_posthoc1[[region]] <- posthoc1
 # write.csv(posthoc1, paste0("Posthoc1_", region, "_Sidak_kmeansclusters.csv"))
  
  # Extract and store posthoc2
  posthoc2 <- stats.testing[[3]]
  posthoc2$correction <- "sidak"
  all_posthoc2[[region]] <- posthoc2
 # write.csv(posthoc2, paste0("Posthoc2_", region, "_Sidak_kmeansclusters.csv"))
}

# Optional: combine and export all results into one table each
combined_anova <- dplyr::bind_rows(all_anovas, .id = "Subregion")
combined_posthoc1 <- dplyr::bind_rows(all_posthoc1, .id = "Subregion")
combined_posthoc2 <- dplyr::bind_rows(all_posthoc2, .id = "Subregion")


#add significance column
combined_anova <- combined_anova  %>%
  mutate(Significance = ifelse(`Pr(>Chisq)` < 0.05, "significant", "ns"))


# Save combined files
write.csv(combined_anova, "Combined_ANOVA_AllSubregions_butyrateonly_ameboidonly.csv")
write.csv(combined_posthoc1, "Combined_Posthoc1_Sidak_AllSubregions_butyrateonly_ameboidonly.csv")
write.csv(combined_posthoc2, "Combined_Posthoc2_Sidak_AllSubregions_butyrateonly_ameboidonly.csv")



```


